Graphical Tests for Power Comparison of Competing Designs

### 2012-03-29 comparison.tex
table formatting
added references, re-worked experiment 1

%% Abstract section.
\abstract{
Lineups  \cite{buja:2009, wickham:2010} have been established as tools for visual testing
similar to standard statistical inference tests, allowing us to
evaluate the validity of graphical findings in an objective manner. In
simulation studies  \cite{majumder:2011} lineups have been shown as being efficient: the
power of visual tests is comparable to classical tests while being
much less stringent in terms of distributional assumptions made. This
makes lineups versatile, yet powerful, tools in situations where
conditions for regular statistical tests are not or cannot be met. In
this paper we introduce lineups as a tool for evaluating the power of
competing graphical designs. We highlight some of the theoretical
properties and then show results from two studies evaluating competing
designs: both studies are designed to go to the threshold of our
perceptual abilities to highlight differences between designs. We use
both accuracy and speed of evaluation as measures of a successful
design. The first study compares the choice of coordinate system:
polar versus Euclidean coordinates. The results show strong support in
favor of Euclidean coordinates in finding fast and accurate answers to
spotting patterns. The second study is aimed at finding shift
differences between distributions. Both studies are motivated by data
problems that we have recently encountered, and explore using
simulated data to evaluate the plot designs under controlled
conditions.  Amazon Turk is used to conduct the studies.  The lineups
provide an effective mechanism for objectively evaluating plot
designs.
} % end of abstract

@@

%% Uncomment below to include a teaser figure.
\teaser{
\centering
\includegraphics[height=1in]{polar}\hspace{0.05\linewidth}
\includegraphics[height=1in]{eucline}
%\includegraphics[width=16cm]{CypressView.eps}
\caption{teaser chart should include all different designs}
}

>>>>

%% Abstract section.
\abstract{
Lineups  \cite{buja:2009, wickham:2010} have been established as tools for visual testing
similar to standard statistical inference tests, allowing us to
evaluate the validity of graphical findings in an objective manner. In
simulation studies  \cite{majumder:2011} lineups have been shown as being efficient: the
power of visual tests is comparable to classical tests while being
much less stringent in terms of distributional assumptions made. This
makes lineups versatile, yet powerful, tools in situations where
conditions for regular statistical tests are not or cannot be met. In
this paper we introduce lineups as a tool for evaluating the power of
competing graphical designs. We highlight some of the theoretical
properties and then show results from two studies evaluating competing
designs: both studies are designed to go to the limits of our
perceptual abilities to highlight differences between designs. We use
both accuracy and speed of evaluation as measures of a successful
design. The first study compares the choice of coordinate system:
polar versus Euclidean coordinates. The results show strong support in
favor of Euclidean coordinates in finding fast and accurate answers to
spotting patterns. The second study is aimed at finding shift
differences between distributions. Both studies are motivated by data
problems that we have recently encountered, and explore using
simulated data to evaluate the plot designs under controlled
conditions.  Amazon Turk is used to conduct the studies.  The lineups
provide an effective mechanism for objectively evaluating plot
designs.
} % end of abstract

@@

%% Uncomment below to include a teaser figure.
\teaser{
\centering
\includegraphics[height=1.2in]{polar}\hspace{0.05\linewidth}
\includegraphics[height=1.2in]{eucline}
%\includegraphics[width=16cm]{CypressView.eps}
\caption{Which of the three plots is different from the others? - Is this question easier to answer with the Polar Charts or the Barcharts?}
}

### 2012-03-29 intro.tex
added references, re-worked experiment 1

Lineups provide a way to quantify the significance of the finding
\citet{majumder:2011}. The significance depends on the size ($m$) of
the lineup ($m=20$ in Figure \ref{lineup}) and the number of
independent observers ($n$). Let $x$ be the number of observers who
correctly identify the plot of the real data. The probability that one
observer correctly identifies the real data plot, in a size $m$
lineup, by chance is $1/m$. To incorporate the results from multiple
observers think about the lineup as $m-1$ head-to-head competitions of
the data plot against $m-1$ null plots, which the data plot has to
win, is picked as the plot with the most distinct features. Assume
that the data plot is the one with the strongest signal, then the
probability that the data plot is picked by $x$ observers follows a
Beta distribution for the maximum \citet{****}:
\[
\text{signal strength} =  \left( \frac{x}{n}\right)^{1/(m-1)}.
\]
We call it signal strength, and it is analgous to power in classical hypothesis testing. Signal strength will range between 0 and 1, with higher values indicating stronger signal, more distinctive feature in the data.

>>>>

Lineups provide a way to quantify the significance of the finding
\citet{majumder:2011}. 
%
The significance depends on the size ($m$) of
the lineup ($m=20$ in Figure \ref{lineup}) and the number of
independent observers ($n$). Let $n$ be the number of independent observers and $x_i$ the number of observers who picked plot $i$, $i \in \{1, ..., m\}$, from the lineup. Then $(x_1, x_2, ..., x_m)$ follows a Multinomial Distribution Mult$_{\pi_1, \pi_2, ..., \pi_m}$ with $\sum_i \pi_i = 1$, where $\pi_i$ is the probability that plot $i$ is picked by an observer, which we can estimate as $\widehat{\pi_i} = x_i/n$.

The power of a test is the probability to reject the null hypothesis. In the framework of lineups, power is the probability to identify the data plot, i.e. power can be estimated as $x/n$, where $x$ is the number of observers who correctly identifies the plot of the real data out of $n$ observers.

%Let $x$ be the number of observers who
%correctly identify the plot of the real data. The probability that one
%observer correctly identifies the real data plot, in a size $m$
%lineup, by chance is $1/m$.

\paragraph{Signal Strength}
To incorporate the results from multiple
observers think about the lineup as $m-1$ head-to-head competitions of
the data plot against $m-1$ null plots, which the data plot has to
win, is picked as the plot with the most distinct features. Assume
that the data plot is the one with the strongest signal, then the
probability that the data plot is picked by $x$ observers follows a
Beta distribution for the maximum \citet{****}:
\[
\text{signal strength} =  \left( \frac{x}{n}\right)^{1/(m-1)}.
\]
We call it signal strength, and it is analogous to 1-$p$-value of a classical hypothesis test. Signal strength will range between 0 and 1, with higher values indicating stronger signal, more distinctive feature in the data.

The significance depends on the size ($m$) of
the lineup ($m=20$ in Figure \ref{lineup}) and the number of
independent observers ($n$). Let $x$ be the number of observers who
correctly identify the plot of the real data. The probability that one
observer correctly identifies the real data plot, in a size $m$
lineup, by chance is $1/m$. To incorporate the results from multiple
observers think about the lineup as $m-1$ head-to-head competitions of
the data plot against $m-1$ null plots, which the data plot has to
win, is picked as the plot with the most distinct features. Assume
that the data plot is the one with the strongest signal, then the
probability that the data plot is picked by $x$ observers follows a
Beta distribution for the maximum \citet{****}:
\[
\text{signal strength} =  \left( \frac{x}{n}\right)^{1/(m-1)}.
\]
We call it signal strength, and it is analgous to power in classical hypothesis testing. Signal strength will range between 0 and 1, with higher values indicating stronger signal, more distinctive feature in the data.

### 2012-03-29 setup.tex
added references, re-worked experiment 1

% !TEX root = comparison.tex
\section{Experimental Setup}~\label{experimentone}
We want two discuss two applications, in which we are making use of lineups to compare competing designs. 
The first one arises from a discussion of how wind direction affects efficiency of an airport. In this example, we are dealing with all flights \cite{rita} in and out of Seattle/Tacoma International Airport (SEA) between July 2008 and June 2011. As a measure of airport efficiency we are using time between successive wheel-events (time at wheel take-off or touch-down), which we found to be independent of airline carrier and operating hours, as long as we restricted the data to `regular' operating hours between 7 am and midnight and `regular' weather conditions -- i.e. we eliminated records associated with the top 5\% wind speed measurements \cite{noaa-weather}, leaving us with just under 0.5 Million flights.

Statistical tests of mean efficiency by wind direction are not particularly helpful in this situation:
the difference in mean efficiency between the most efficient wind direction and the least efficient shows a statistically significant difference with a $p$-value $\le 10^{15}$. %2.2e-16$. 
However, out of the 34 other wind directions,  another 31 show significant differences in efficiency as well. This is much more a property of the large data size rather than practically usable differences. Mere significances also do not allow us an assessment of the underlying pattern.

\subsection{Wind Direction and Airport Efficiency}
In deciding on the design for displaying efficiency  by wind direction we were using the fact that wind direction is circular, and  displayed the data as (conditional) wind rose charts - i.e. for each of 36 wind directions we show the percentage of flights falling into one-minute intervals between successive flights, from  0 minutes to 8 minutes or more.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
%   \centering
 \hspace{-.1in}
   \begin{tabular}{ccl}
   \includegraphics[width=0.375\linewidth]{euclid_noline.pdf} & \hspace{-.3in}
   \includegraphics[width=0.375\linewidth]{euclid_line.pdf} &  \hspace{-.2in} \multirow{2}{*}[.55in]{  \includegraphics[width=0.2\linewidth]{legend.pdf}} \\
   \includegraphics[width=0.43\linewidth]{Polar_NoLine.pdf} &  \hspace{-.3in}
   \includegraphics[width=0.43\linewidth]{Polar_Line.pdf}  
   \end{tabular}
   \caption{All four competing designs: polar versus Euclidean charts in top and bottom row, with (right) and without (left) reference lines at 50\%. }
   \label{layouts}
\end{figure}

@@

An additional factor we were interested in this particular situation was to assess, how much of the data we actually needed to use in a design to have observers pick out the pattern. Clearly, a design is more efficient, if a smaller sample size is sufficient for showing the presence of a relationship. In order to investigate this, we took different size samples of the data and created lineup plots of all four designs for these subsets. The effect of sampling size on the displays mostly stems from the additional variability introduced when using small samples, which might hide the pattern, if it is not displayed prominently. 
XXX picture of sampling size?
Another perturbation to the data are `shifts' in wind direction, i.e. we make use of the circular nature of the wind direction and adjust the $x$ axis by using different offsets.

@@

 This results in a shift of the wave in Euclidean charts and a rotation in the polar charts as can be seen in figure \ref{fig:offset}. Our initial hypothesis was that this would have no effect on the polar charts, but might have a deteriorating effect on the Euclidean charts, in the case that the peak or the valley of the wave was at either end of the $x$ axis. 
While these simple perturbations and different sample sizes allow us to get insight into different aspects of the designs, they also allow us to get multiple responses from each observer to assess individual ability without the need to go outside the framework of the original data, i.e. we leave the joint relationship between $x$ and $y$ essentially unchanged. For all of these combinations we produced two replicates, resulting in a total of 192 different lineups ($ 4 \text{ designs} \times 6 \text{ sample sizes} \times 4 \text{ offsets} \times 2 \text{ reps})$.

Data was collected using Amazon's Mechanical Turk (MTurk) service. XXX More details: website and screen shot.
 Each participant was shown ten lineups and asked to choose the graphic on each chart which they think is ``different''. They then identify, from a list of choices, why they chose this particular chart as well as how confident they are on a scale of one to five. Personal information such as age group, gender, and education is collected on a voluntary basis. 
The amount of time it took the individual to answer is also recorded. This study allows us to assess how design choices affect decoding \citet{cleveland:1994} in terms of accuracy and speed.

@@

%Besides correctness of responses, two additional response variables were recorded:  time taken for answer and a (subjective) confidence level, ranging from 1 (lowest) to 5 (highest), assessing how convinced participants are of the correctness of their answer. 
Figure \ref{fig:time} shows time taken until a decision was made in histograms, facetted by design and colored by correctness of answer. because of the skewness of the distributions, times were log-transformed. Polar charts take on average more time to answer, but are answered with much lower precision. Figure \ref{fig:conf} shows two barcharts of confidence levels by task, again coloring is used for correctness of answers. Euclidean displays lead to a very bimodal distribution of confidence: participants are either very sure or not sure at all of their answer. Confidence levels in polar charts are distributed much more uniformly.

@@

Most generally this experiment can be paralleled to the pie chart versus barchart discussion \cite{***}. 
The Euclidean coordinate plot is made up of bars where areas are colored proportionally according to percentages. The polar coordinate plot is made up of pie slices which are also colored proportionally according to the data. 
Viewers of these charts will visually decode the areas of the bars and pie slices and make a judgement based on their decoding. According to  \citet[page=40]{kosslyn:2006}, area is usually perceived as a function of what the actual area is. This function is the area raised to an exponent of approximately 0.8 and then multiplied by a constant. However, when bars are parallel, the relative height of these bars is perceived very accurately. From this information, one could infer that the Euclidean coordinate plot may gain efficiency based on the height of the colored bars.

@@

\begin{table}[ht]
\begin{center}
\resizebox{\linewidth}{!} {
\rowcolors{2}{white}{lightgray}
\begin{tabular}{rrrrrrl}
  \hline
& & Estimate & Error & $z$-value & $p$-value &\\ 
  \hline
design & euclid & -0.08 & 0.39 & -0.21 & 0.84 &\\ 
&polar & -1.98 & 0.32 & -6.13 & 0.00 & ***\\ [3pt]
main effects &&&&&&\\
reference & line  & -0.14 & 0.26 & -0.53 & 0.59 & \\ [1pt]
size &  sample size & 0.27 & 0.04 & 6.31 & 0.00 & ***\\[1pt]
offset &  90 degrees& -0.43 & 0.37 & -1.18 & 0.24 &\\ 
  & 180 degrees& -0.89 & 0.35 & -2.51 & 0.01 & **\\ 
  & 270 degrees& 0.21 & 0.38 & 0.55 & 0.58 &\\ [3pt]
interactions  &&&&&&\\
&  polar:line & 0.51 & 0.35 & 1.44 & 0.15 &\\ [1pt]
&    polar:sample size & -0.23 & 0.05 & -5.02 & 0.00 & ***\\[1pt]
&    polar:offset 90 & 0.64 & 0.49 & 1.30 & 0.20 \\ 
&  polar:offset 180 & 0.91 & 0.47 & 1.92 & 0.05 & .\\ 
&    polar:offset 270 & -0.73 & 0.54 & -1.35 & 0.18 &\\
   \hline
\end{tabular}
}
\end{center}
\caption{\label{tbl:correct} Output of a generalized linear mixed effects model for power of lineups (i.e. probability of identifying the data plot) for comparing designs. Included are two-way interactions with sample size and shifts in wind direction (offset). Results are based on  976 lineup evaluations by 115 participants. }
\end{table}

>>>>

% !TEX root = comparison.tex
\section{Experimental Setup}~\label{experimentone}
We want two discuss two applications, in which we are making use of lineups to compare competing designs. 
The first one arises from a discussion of how wind direction affects efficiency of an airport. in the second example we are interested in visualizing mean shifts between two distributions.

\subsection{Wind Direction and Airport Efficiency}
In this example, we are dealing with all flights \cite{rita} in and out of Seattle/Tacoma International Airport (SEA) between July 2008 and June 2011. As a measure of airport efficiency we are using time between successive wheel-events (time at wheel take-off or touch-down), which we found to be independent of airline carrier and operating hours, as long as we restricted the data to `regular' operating hours between 7 am and midnight and `regular' weather conditions -- i.e. we eliminated records associated with the top 5\% wind speed measurements \cite{noaa-weather}, leaving us with just under 0.5 Million flights.

Statistical tests of mean efficiency by wind direction are not particularly helpful in this situation:
the difference in mean efficiency between the  wind direction at which the airport operates the most efficiently and its least efficient direction shows a statistically significant difference with a $p$-value $\le 10^{15}$. %2.2e-16$. 
However, out of the 34 other wind directions,  another 31 show significant differences in efficiency as well. This is much more a property of the large data size rather than practically usable differences. Mere significances also do not allow us an assessment of the underlying pattern.

In deciding on the design for displaying efficiency  by wind direction we were using the fact that wind direction is circular, and  displayed the data as (conditional) wind rose charts - i.e. for each of 36 wind directions we show the percentage of flights falling into one-minute intervals between successive flights, from  0 minutes to 8 minutes or more.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
%   \centering
 \hspace{-.1in}
   \begin{tabular}{ccl}
   \includegraphics[width=0.43\linewidth]{Polar_NoLine.pdf} &  \hspace{-.3in}
   \includegraphics[width=0.43\linewidth]{Polar_Line.pdf}  &  \hspace{-.2in} \multirow{2}{*}[.55in]{  \includegraphics[width=0.2\linewidth]{legend.pdf}} \\
   \includegraphics[width=0.375\linewidth]{euclid_noline.pdf} & \hspace{-.3in}
   \includegraphics[width=0.375\linewidth]{euclid_line.pdf}
   \end{tabular}
   \caption{All four competing designs: polar versus Euclidean charts in top and bottom row, with (right) and without (left) reference lines at 50\%. }
   \label{layouts}
\end{figure}

@@

An additional factor we were interested in this particular situation was to assess, how much of the data we actually needed to use in a design to have observers pick out the pattern. Clearly, a design is more efficient, if a smaller sample size is sufficient for showing the presence of a relationship. In order to investigate this, we took different size samples of the data and created lineup plots of all four designs for these subsets. The effect of sampling size on the displays mostly stems from the additional variability introduced when using small samples, which might hide the pattern, if it is not displayed prominently. 
%XXX picture of sampling size?
Another perturbation to the data are `shifts' in wind direction, i.e. we make use of the circular nature of the wind direction and adjust the $x$ axis by using different offsets.

@@

 This results in a shift of the wave in Euclidean charts and a rotation in the polar charts as can be seen in figure \ref{fig:offset}. Our initial hypothesis was that this would have no effect on the polar charts, but might have a deteriorating effect on the Euclidean charts, in the case that the peak or the valley of the wave was at either end of the $x$ axis, i.e. an offset of 90 or 270 degrees. 
While these simple perturbations and different sample sizes allow us to get insight into different aspects of the designs, they also allow us to get multiple responses from each observer to assess individual ability without the need to go outside the framework of the original data, i.e. we leave the joint relationship between $x$ and $y$ essentially unchanged. For all of these combinations we produced two replicates, resulting in a total of 192 different lineups ($ 4 \text{ designs} \times 6 \text{ sample sizes} \times 4 \text{ offsets} \times 2 \text{ reps})$.

Data was collected using Amazon's Mechanical Turk (MTurk) service. XXX More details: website and screen shot.
 Each participant was shown ten lineups and asked to choose the graphic on each chart which they think is ``different''. They then identify, from a list of choices, why they chose this particular chart as well as how confident they are on a scale of one to five. Personal information such as age group, gender, and education is collected on a voluntary basis. 
The amount of time it took the individual to answer is also recorded. This  allows us to assess how design choices affect decoding \citet{cleveland:1994} in terms of accuracy and speed.

@@

%Besides correctness of responses, two additional response variables were recorded:  time taken for answer and a (subjective) confidence level, ranging from 1 (lowest) to 5 (highest), assessing how convinced participants are of the correctness of their answer. 
Figure \ref{fig:time} shows time taken until a decision was made in histograms, facetted by design and colored by correctness of answer. Because of the skewness of the distributions, times were log-transformed. Polar charts take on average more time to answer, but are answered with much lower precision. Figure \ref{fig:conf} shows two barcharts of confidence levels by task, again coloring is used for correctness of answers. Euclidean displays lead to a very bimodal distribution of confidence: participants are either very sure or not sure at all of their answer. Confidence levels in polar charts are distributed much more uniformly.

@@

The perceptual tasks involved in decoding the designs fit into the general framework of the piechart versus barchart discussion. However, we are not dealing with assessing angles, which is known to be a perceptual harder task than assessments of heights or widths in bars, see e.g. \cite{cleveland:1984, robbins:2004, few:2009}. Instead, the perceptual tasks consists of comparisons along a common axis (in the barcharts) and a common origin (in the polar charts). The difference in designs is therefore based on how well we are able to judge deviations from a horizontal line compared to deviations from a circle. Based on this, the charts with added reference lines provide us with exactly the frame we compare with and should, therefore, be the `better' designs - either in speed or accuracy.

%Most generally this experiment can be paralleled to the pie chart versus barchart discussion \cite{***}. 
%The Euclidean coordinate plot is made up of bars where areas are colored proportionally according to percentages. The polar coordinate plot is made up of pie slices which are also colored proportionally according to the data. 
%Viewers of these charts will visually decode the areas of the bars and pie slices and make a judgement based on their decoding. According to  \citet[page=40]{kosslyn:2006}, area is usually perceived as a function of what the actual area is. This function is the area raised to an exponent of approximately 0.8 and then multiplied by a constant. However, when bars are parallel, the relative height of these bars is perceived very accurately. From this information, one could infer that the Euclidean coordinate plot may gain efficiency based on the height of the colored bars.

@@

\begin{table}[ht]
\begin{center}
\resizebox{\linewidth}{!} {
%\rowcolors{2}{white}{lightgray}
\begin{tabular}{rrrrrrl}
  \hline
& & Estimate & Error & $z$-value & $p$-value &\\ 
  \hline
\bf design & euclid & -0.08 & 0.39 & -0.21 & 0.84 &\\ 
&polar & -1.98 & 0.32 & -6.13 & 0.00 & ***\\ [3pt]
\multicolumn{2}{l}{\bf main effects} &&&&&\\
& reference line  & -0.14 & 0.26 & -0.53 & 0.59 & \\ [1pt]
&  sample size & 0.27 & 0.04 & 6.31 & 0.00 & ***\\[1pt]
 &offset:  90 degrees& -0.43 & 0.37 & -1.18 & 0.24 &\\ 
  & 180 degrees& -0.89 & 0.35 & -2.51 & 0.01 & **\\ 
  & 270 degrees& 0.21 & 0.38 & 0.55 & 0.58 &\\ [3pt]
\multicolumn{2}{l}{\bf interactions} &&&&&\\
&  polar:line & 0.51 & 0.35 & 1.44 & 0.15 &\\ [1pt]
&    polar:sample size & -0.23 & 0.05 & -5.02 & 0.00 & ***\\[1pt]
&    polar:offset 90 & 0.64 & 0.49 & 1.30 & 0.20 \\ 
&  polar:offset 180 & 0.91 & 0.47 & 1.92 & 0.05 & .\\ 
&    polar:offset 270 & -0.73 & 0.54 & -1.35 & 0.18 &\\
   \hline
\end{tabular}
}
\end{center}
\caption{\label{tbl:correct} Output of a generalized linear mixed effects model for power of lineups (i.e. probability of identifying the data plot) for comparing designs. Included are two-way interactions with sample size and shifts in wind direction (offset). Results are based on  976 lineup evaluations by 115 participants. }
\end{table}
